---
title: "testing_gp_models"
output: html_document
---

After running the first round of emulator fitting and MCMC sampling, the 
stationary GP models used appeared not to be appropriate for this 
experiment. This document explores alternative combinations of models and
initial designs that are more appropriate. The motivating idea is to better
capture the global trend to produce GP priors that are reasonable with respect
to the goal of modeling probability density functions. We try to accomplish
this by combining three strategies:
1.) Modeling the log unnormalized posterior density, rather than the log-likelihood,
so that the effect of the prior driving down the tails can be leveraged by
the GP model.
2.) Fitting a trend that decreases as distance from the design increases. For now
we consider quadratic trends, but more sophisticated approaches could be 
considered (e.g., Roshan Joseph's composite GP approach).
3.) Choose an initial design geared towards both capturing the global trend,
as well as local variation. A simple point is to use a space-filling design
within the bulk of the prior distribution, then sample some points way
out in the tails.

There are different considerations depending on whether forward model 
emulation or log density emulation is utilized.

Forward model: in general, the behavior of the forward model in the tails
of the prior is not known. There is no distinction analogous to 
"log-likelihood vs. log-unnormalized posterior density" in this setting.
However, there still is opportunity for better initial designs. My hypothesis
is that including some extrapalatory initial design points can still help deal
with the tail issue. This should probably be combined with a global trend term, 
but there is less prior information on what this trend should look like as 
opposed to the log-unnormalized posterior density case. This seems like a
good case for something like Roshan Joseph's composite GP approach.

Log density: In addition to the design considerations, this setting has the 
additional question of whether to emulate the log-likelihood or log unnormalied
posterior density. This distinction may seem unimportant; given a
log-likelihood emulator, a log-posterior emulator can be constructed by simply
adding the log prior. It seems to me, however, that the potential benefit 
of the log-posterior emulation approach is that in this setting we have 
more (at least qualitative) prior information about what the black-box 
function should look like. The log-likelihood can be quite flat in certain
regions, and may not be decreasing even far in the tails of the prior.
However, with the log-posterior we know the tails have to decrease and can 
make this information into the GP model using a global trend.


```{r}
set.seed(543653)

library(data.table)
library(ggplot2)
library(patchwork)
library(scales)
library(grid)
library(support)


experiment_tag <- "banana"

base_dir <- file.path("/projectnb", "dietzelab", "arober", "bip-surrogates-paper")
code_dir <- file.path("/projectnb", "dietzelab", "arober", "gp-calibration")
src_dir <- file.path(code_dir, "src")
experiment_dir <- file.path(base_dir, "experiments", experiment_tag)
inv_prob_dir <- file.path(experiment_dir, "output", "inv_prob_setup")
settings_dir <- file.path(experiment_dir, "output", "alg_settings")

# Source required files.
source(file.path(src_dir, "general_helper_functions.r"))
source(file.path(src_dir, "statistical_helper_functions.r"))
source(file.path(src_dir, "plotting_helper_functions.r"))
source(file.path(src_dir, "mcmc_helper_functions.r"))
source(file.path(src_dir, "seq_design.r"))
source(file.path(src_dir, "gp_helper_functions.r"))
source(file.path(src_dir, "gpWrapper.r"))
source(file.path(src_dir, "llikEmulator.r"))
source(file.path(src_dir, "gp_mcmc_functions.r"))
source(file.path(base_dir, "scripts", "helper", "sim_study_functions.r"))
```

# Read inputs for analysis

```{r}
# Inverse problem object and algorithm/model settings.
inv_prob <- readRDS(file.path(inv_prob_dir, "inv_prob_list.rds"))
em_settings <- readRDS(file.path(settings_dir, "em_settings.rds"))
mcmc_settings <- readRDS(file.path(settings_dir, "mcmc_settings.rds"))
```

```{r}
# Read exact MCMC samples.
samp_dt <- fread(file.path(out_dir, "samp_exact.csv"))
test_info_prior <- readRDS(file.path(out_dir, "test_info_prior.rds"))
test_info_post <- readRDS(file.path(out_dir, "test_info_post.rds"))
```

# Set up grid for plotting
```{r}
# 2d grid for plotting. Include extrapolations well into the tails of the prior.
# Note that priors in this case are independent N(0,1) for each parameter.
q <- qnorm(c(.001, .999))
extrap_dist <- c(-2,2)

input_grid <- get_batch_design("tensor_product_grid", N_batch=50^2,
                               bounds=cbind(q+extrap_dist,q+extrap_dist))
colnames(input_grid) <- inv_prob$par_names
test_info_grid <- get_init_design_list(inv_prob, inputs=input_grid, 
                                       design_method="tensor_product_grid",
                                       N_design=50^2)

# Data for plotting.
plt_df <- data.table(test_info_grid$input, lprior=test_info_grid$lprior,
                     llik=test_info_grid$llik, lpost=test_info_grid$lpost,
                     fwd=drop(test_info_grid$fwd))
plt_df$post <- exp(plt_df$lpost)

# Subsample from posterior.
post_samp_subset <- as.data.frame(select_mcmc_samp_mat(samp_dt, thin=100L))

# For drawing box that captures more prior mass. These are the corners of 
# the box (must be ordered sequentially - here, clockwise).
input_prior_bounds <- rbind(q, c(q[2], q[2]), -q, c(q[1], q[1]))
colnames(input_prior_bounds) <- inv_prob$par_names
input_prior_bounds <- as.data.table(input_prior_bounds)
```

```{r}
# Base plots for forward model, log prior, log likelihood, log posterior,
# and posterior density.
plt_fwd <- ggplot() +
              geom_raster(aes(x=u1, y=u2, fill=fwd), plt_df, interpolate=TRUE) +
              scale_fill_viridis_c(name="fwd") +
              coord_fixed() +
              ggtitle("Forward model") + 
              geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") +
              theme_minimal()
plt_lprior <- ggplot() +
               geom_contour_filled(aes(x=u1, y=u2, z=lprior), plt_df) +
               ggtitle("Log Prior Density") +
               geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white")
plt_llik <- ggplot() +
              geom_raster(aes(x=u1, y=u2, fill=llik), plt_df, interpolate=TRUE) +
              scale_fill_viridis_c(name="llik") +
              coord_fixed() +
              ggtitle("Log Likelihood") + 
              geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white")
plt_post <- ggplot() +
              geom_raster(aes(x=u1, y=u2, fill=exp(lpost)), plt_df, interpolate=TRUE) +
              scale_fill_viridis_c(name="Post") +
              coord_fixed() +
              ggtitle("Posterior density") + 
              geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white")
plt_lpost <- ggplot() +
              geom_contour_filled(aes(x=u1, y=u2, z=lpost), plt_df) +
              ggtitle("Log Posterior density") +
              geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white")

plot(plt_fwd)
plot(plt_lprior)
plot(plt_llik)
plot(plt_lpost)
plot(plt_post)
```

# Initial Designs.

```{r}
#
# Space-filling design, no extrapolation points.
#

n_design <- 7L
design_info_1 <- get_init_design_list(inv_prob, design_method="LHS", N_design=n_design)
design_info_dt_1 <- data.table(design_info_1$input,
                               fwd = drop(design_info_1$fwd),
                               llik = design_info_1$llik,
                               lprior = design_info_1$lprior,
                               lpost = design_info_1$lpost)
```


```{r}
#
# Design with extrapolation points.
#

# Number of non-extrapolation points.
n_internal <- 3L

# "Internal" design points.
design_internal <- get_init_design_list(inv_prob, design_method="LHS", N_design=n_internal)

# "Extrapolation" design points. Currently setting as a percentage of the 
# prior quantiles.
extrap_scale <- 1.8
extrap_inputs <- rbind(c(0, extrap_scale * q[2]),
                       c(extrap_scale * q[2], 0),
                       c(0, extrap_scale * q[1]),
                       c(extrap_scale * q[1], 0))
colnames(extrap_inputs) <- inv_prob$par_names
design_extrap <- get_init_design_list(inv_prob, "manual", 4, inputs=extrap_inputs)

# Combine design.
design_info_2 <- list(input = rbind(design_internal$input, design_extrap$input),
                    fwd = rbind(design_internal$fwd, design_extrap$fwd),
                    llik = c(design_internal$llik, design_extrap$llik),
                    lprior = c(design_internal$lprior, design_extrap$lprior),
                    lpost = c(design_internal$lpost, design_extrap$lpost))

design_info_dt_2 <- data.table(design_info_2$input,
                             fwd = drop(design_info_2$fwd),
                             llik = design_info_2$llik,
                             lprior = design_info_2$lprior,
                             lpost = design_info_2$lpost)
```


# Forward model emulators.
```{r}
# Fit stationary GPs.
em_fwd_1 <- em_settings$em_fwd$fit_em(design_info_1, inv_prob)
em_fwd_2 <- em_settings$em_fwd$fit_em(design_info_2, inv_prob)

# Forward model predictions.
pred_fwd_1 <- em_fwd_1$predict_emulator(test_info_grid$input)
pred_fwd_2 <- em_fwd_2$predict_emulator(test_info_grid$input)
plt_df[, `:=`(mean_fwd_1=drop(pred_fwd_1$mean), var_fwd_1=drop(pred_fwd_1$var),
              mean_fwd_2=drop(pred_fwd_2$mean), var_fwd_2=drop(pred_fwd_2$var))]

# Log-likelihood predictions.
llik_pred_fwd_1 <- em_fwd_1$predict(test_info_grid$input, em_pred_list=pred_fwd_1)
llik_pred_fwd_2 <- em_fwd_2$predict(test_info_grid$input, em_pred_list=pred_fwd_2)
plt_df[, `:=`(llik_mean_fwd_1=drop(llik_pred_fwd_1$mean), llik_var_fwd_1=drop(llik_pred_fwd_1$var),
              llik_mean_fwd_2=drop(llik_pred_fwd_2$mean), llik_var_fwd_2=drop(llik_pred_fwd_2$var))]

# pred_df_long<- data.table::melt(plt_df, id.vars=c("u1", "u2"), 
#                         variable.name="quantity", value.name="value")
```

## Forward model predictive distribution

### Design 1
```{r}
plt_mean_fwd_1 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=mean_fwd_1), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("Fwd mean pred")

plt_sd_fwd_1 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=sqrt(var_fwd_1)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="sd") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("Fwd sd pred")

plot(plt_fwd + plt_mean_fwd_1 + plt_sd_fwd_1)
```

### Design 2
```{r}
plt_mean_fwd_2 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=mean_fwd_2), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_2, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("Fwd mean pred")

plt_sd_fwd_2 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=sqrt(var_fwd_2)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="sd") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_2, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("Fwd sd pred")

plot(plt_fwd + plt_mean_fwd_2 + plt_sd_fwd_2)
```

## Induced log-likelihood predictive distribution

### Design 1

```{r}
plt_llik_mean_fwd_1 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=llik_mean_fwd_1), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("llik mean pred")

plt_llik_sd_fwd_1 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=sqrt(llik_var_fwd_1)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="sd") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("llik sd pred")

plot(plt_llik + plt_llik_mean_fwd_1 + plt_llik_sd_fwd_1)
```

```{r}
em_fwd_1$plot_pred_validation(test_info_grid$input, em_pred_list=pred_fwd_1,
                              true_llik=test_info_grid$llik)
```


```{r}
plt_lik_mean_fwd_1 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=exp(llik_mean_fwd_1)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), post_samp_subset, color="gray", alpha=0.5) + 
                  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("lik mean pred")
plot(plt_lik_mean_fwd_1)
```


### Design 2

```{r}
plt_llik_mean_fwd_2 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=llik_mean_fwd_2), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_2, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("llik mean pred")

plt_llik_sd_fwd_2 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=sqrt(llik_var_fwd_2)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="sd") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), design_info_dt_2, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("llik sd pred")

plot(plt_llik + plt_llik_mean_fwd_2 + plt_llik_sd_fwd_2)
```

```{r}
plt_lik_mean_fwd_2 <- ggplot() + 
                  geom_raster(aes(x=u1, y=u2, fill=exp(llik_mean_fwd_2)), plt_df, interpolate=TRUE) +
                  scale_fill_viridis_c(name="mean") +
                  coord_fixed() +
                  geom_point(aes(x=u1, y=u2), post_samp_subset, color="gray", alpha=0.5) + 
                  geom_point(aes(x=u1, y=u2), design_info_dt_2, color="red") + 
                  geom_polygon(data=input_prior_bounds, aes(x=u1, y=u2), fill=NA, color="white") + 
                  ggtitle("lik mean pred")
plot(plt_lik_mean_fwd_2)
```



```{r}
em_fwd_2$plot_pred_validation(test_info_grid$input, em_pred_list=pred_fwd_2,
                              true_llik=test_info_grid$llik)
```

## MCMC: forward model emulator design 1
```{r}
mcmc_settings_fwd <- mcmc_settings[c("mean", "marginal", "mcwmh-ind", "mcwmh-joint")]

mcmc_results_fwd_1 <- run_mcmc_comparison(em_fwd_1, inv_prob$par_prior, 
                                          mcmc_settings_fwd, return=TRUE)
```

```{r}
mcmc_mean_fwd_1 <- mcmc_results_fwd_1$mean$samp
trace_plts <- get_trace_plots(mcmc_mean_fwd_1)
trace_plt_llik <- get_trace_plots(mcmc_results_fwd_1$mean$info,
                                  param_types="dens", param_names="llik")

for(plt in trace_plts) plot(plt)
plot(trace_plt_llik[[1]])

mcmc_mean_fwd_1 <- select_mcmc_samp_mat(mcmc_mean_fwd_1, thin=10)
support_pts <- support::sp(n=2000L, p=2, dist.samp=mcmc_mean_fwd_1)
mcmc_mean_fwd_1 <- support_pts$sp
colnames(mcmc_mean_fwd_1) <- inv_prob$par_names
```

```{r}
plt_post + 
  geom_point(aes(x=u1, y=u2), design_info_dt_1, color="red") + 
  geom_point(aes(x=u1, y=u2), as.data.table(mcmc_mean_fwd_1), color="gray", alpha=0.5)
```

## MCMC: forward model emulator design 2
```{r}
mcmc_results_fwd_2 <- run_mcmc_comparison(em_fwd_2, inv_prob$par_prior, 
                                          mcmc_settings_fwd, return=TRUE)
```
      


# Log-likelihood/Log-posterior emulators.
```{r}
# Fit GP for log-likelihood.
gp_obj <- gpWrapperHet(design_info$input, matrix(design_info$llik, ncol=1),
                       scale_input=TRUE, normalize_output=TRUE)
gp_obj$set_gp_prior("Gaussian", "constant", include_noise=FALSE)
gp_obj$fit()

# Instantiate and save log-likelihood emulator object.
llik_em <- llikEmulatorGP("em_llik", gp_obj, default_conditional=FALSE, 
                          default_normalize=TRUE, 
                          lik_par=inv_prob$llik_obj$get_lik_par(), 
                          llik_bounds=inv_prob$llik_obj$get_llik_bounds())



```






