---
title: "noisy_mcmc_1d_noise_lvl_test"
output: html_document
---

Tests to do:
- Compare algs when noise level of GP is varied.
- Effect of stationary prior; test what happens with noisy MCMC.

```{r}
set.seed(63462)

library(data.table)
library(ggplot2)
library(patchwork)
library(scales)
library(grid)
library(support)

base_dir <- file.path("/projectnb", "dietzelab", "arober", "bip-surrogates-paper")
code_dir <- file.path("/projectnb", "dietzelab", "arober", "gp-calibration")
src_dir <- file.path(code_dir, "src")

# Source required files.
source(file.path(src_dir, "general_helper_functions.r"))
source(file.path(src_dir, "statistical_helper_functions.r"))
source(file.path(src_dir, "plotting_helper_functions.r"))
source(file.path(src_dir, "mcmc_helper_functions.r"))
source(file.path(src_dir, "seq_design.r"))
source(file.path(src_dir, "gp_helper_functions.r"))
source(file.path(src_dir, "gpWrapper.r"))
source(file.path(src_dir, "llikEmulator.r"))
source(file.path(base_dir, "scripts", "helper", "sim_study_functions.r"))
```

# Inverse problem setup
```{r}

# TODO:
# Maybe do something more grounded in a real problem here. Estimating
# spring constant? Or parameter in Volterra model?

# Identity forward model.
fwd <- function(x) 3*x

# Gaussian likelihood.
llik <- function(x) dnorm(fwd(x), mean=0, sd=1, log=TRUE)

# Gaussian prior.
prior_var <- 1
par_prior <- data.frame(dist="Gaussian", param1=0, param2=sqrt(prior_var))
lprior <- get_lprior_dens(par_prior)
```

# Grid for plotting.
```{r}
n_grid <- 100L
grid_min <- -8
grid_max <- 8

grid_input <- seq(grid_min, grid_max, length.out=n_grid)
design_grid <- list(input=cbind(u=grid_input), fwd=cbind(y=fwd(grid_input)),
                    llik=llik(grid_input))
design_grid_dt <- data.table(u=drop(design_grid$input), fwd=drop(design_grid$fwd),
                             llik=design_grid$llik, lprior=lprior(design_grid$input))
design_grid_dt[, lpost := llik + lprior]
```

```{r}
plt_fwd <- ggplot() +
           geom_line(aes(x=u, y=fwd), design_grid_dt, color="black")

plt_dens <- ggplot() + 
              geom_line(aes(x=u, y=llik), design_grid_dt, color="green") +
              geom_line(aes(x=u, y=lprior), design_grid_dt, color="blue") +
              geom_line(aes(x=u, y=lpost), design_grid_dt, color="red") +
              ylim(-30, 0)

plot(plt_fwd)
plot(plt_dens)
```






# Design
```{r}
n_design <- 4L
design_input <- seq(prior_min, prior_max, length.out=n_design)
design_info <- list(input=cbind(u=design_input), fwd=cbind(y=fwd(design_input)),
                    llik=llik(design_input))
design_info_dt <- data.table(u=drop(design_info$input), fwd=drop(design_info$fwd),
                             llik=design_info$llik)
```

```{r}
plt_fwd <- ggplot() +
           geom_line(aes(x=u, y=fwd), design_grid_dt, color="black") +
           geom_point(aes(x=u, y=fwd), design_info_dt, color="red")

plt_llik <- ggplot() + 
            geom_line(aes(x=u, y=llik), design_grid_dt, color="black") + 
            geom_point(aes(x=u, y=llik), design_info_dt, color="red")

plot(plt_fwd)
plot(plt_llik)
```

# Fit forward model and log-likelihood GPs.
```{r}
# GPs.
gp_fwd <- gpWrapperHet(design_info$input, design_info$fwd,
                       scale_input=TRUE, normalize_output=TRUE)
gp_fwd$set_gp_prior("Gaussian", "constant", include_noise=FALSE)
gp_fwd$fit()

# Log-likelihood emulators.
gp_llik <- gpWrapperHet(design_info$input, matrix(design_info$llik, ncol=1L),
                        scale_input=TRUE, normalize_output=TRUE)
gp_llik$set_gp_prior("Gaussian", "constant", include_noise=FALSE)
gp_llik$fit()
```

Plan: vary the noise level in the GPs by simply scaling the GP variance 
prediction. Run noisy MCMC algs at different levels. Compare each to
pseudo-marginal and closed-form marginal. Do we see agreement between
pseudo-marginal/marginal and mcwmh-ind at small noise levels? How do the 
algorithms diverge at large noise levels?












