---
title: "noisy_mcmc_1d_noise_lvl_test"
output: html_document
---

Tests to do:
- Compare algs when noise level of GP is varied.
- Effect of stationary prior; test what happens with noisy MCMC.

Nice examples:

List 1:

List 2:
#2: Example where marginal is multimodal but mcwmh-joint is not (results in 
basically inflating the tails). mcwmh-ind has some divergence issues.
Design points: (-7, -2, 2, 7)

#6: This is an interesting one that illustrates a few different points.
Shows the benefit of the marginal approx over the mean approx. Interesting case
where the marginal sampler converges but the mcwmh samplers diverge. This makes
complete sense in this example: 



```{r}
set.seed(63462)

library(data.table)
library(ggplot2)
library(patchwork)
library(scales)
library(grid)
library(support)

base_dir <- file.path("/Users/andrewroberts/Desktop/git-repos/bip-surrogates-paper")
code_dir <- file.path("/Users/andrewroberts/Desktop/git-repos/gp-calibration")
src_dir <- file.path(code_dir, "src")

# Source required files.
source(file.path(src_dir, "general_helper_functions.r"))
source(file.path(src_dir, "statistical_helper_functions.r"))
source(file.path(src_dir, "plotting_helper_functions.r"))
source(file.path(src_dir, "mcmc_helper_functions.r"))
source(file.path(src_dir, "seq_design.r"))
source(file.path(src_dir, "gp_helper_functions.r"))
source(file.path(src_dir, "gpWrapper.r"))
source(file.path(src_dir, "llikEmulator.r"))
source(file.path(base_dir, "scripts", "helper", "sim_study_functions.r"))
```


# Bayesian inverse problem setup.
```{r}
# Forward model.
fwd <- function(U) {
  if(is.null(dim(U))) U <- matrix(U, ncol=1L)
  colnames(U) <- "u"
  return(U)
}

# Prior.
sig2_prior <- 1
par_prior <- data.frame(dist="Gaussian", param1=0, param2=sig2_prior, lower=-Inf, upper=-Inf)
rownames(par_prior) <- "u"
lprior <- function(U) drop(dnorm(U, mean=0, sd=sqrt(sig2_prior), log=TRUE))

# Ground truth.
u_true <- 0.3
fwd_true <- fwd(u_true)

# Observed data.
n_obs <- 100L
sig2 <- 1
y <- rnorm(n=n_obs, fwd_true, sqrt(sig2))

# Log-likelihood and posterior.
llik <- function(U) {
  if(is.null(dim(U))) U <- matrix(U, ncol=1L)
  f <- function(u) sum(dnorm(y, fwd(u), sqrt(sig2), log=TRUE))
  apply(U, 1, f)
}

lpost <- function(U) llik(U) + lprior(U)
```


# Helper functions.

```{r}
get_design <- function(U) {
  design <- list(input = U)
  design$llik <- llik(design$input)
  design$lprior <- lprior(design$input)
  design$lpost <- lpost(design$input)
  design <- as.data.table(design)
  
  return(design)
}

plot_lpost_em <- function(lpost_em, grid_info, design_info) {
    # Log-posterior surrogate predictions.
  U <- matrix(grid_info$input, ncol=1L)
  pred <- lpost_em$predict(U)
  q_lower <- lpost_em$calc_quantiles(U, p=0.01)
  q_upper <- lpost_em$calc_quantiles(U, p=0.99)

  dt_pred <- data.table(input = grid_info$input,
                        mean = pred$mean,
                        lower = q_lower,
                        upper = q_upper,
                        true = grid_info$lpost)

  # Design used to fit emulator.
  dt_design <- as.data.table(design_info)


  plt <- ggplot(dt_pred, aes(x=input)) +
            geom_line(aes(y=mean), color="blue") +
            geom_ribbon(aes(ymin=lower, ymax=upper), 
                        fill="skyblue", alpha=0.4) +
            geom_line(aes(y=true), color="black") +
            geom_point(aes(x=input, y=lpost), dt_design, 
                       color="red", inherit.aes=FALSE) +
            labs(xlab="u", ylab="log-posterior")
}


plot_post_em <- function(lpost_em, grid_info, design_info) {
  # Log-posterior surrogate predictions.
  U <- matrix(grid_info$input, ncol=1L)
  pred <- lpost_em$predict_lik(U, log_scale=TRUE)
  q_lower <- lpost_em$calc_lik_quantiles(U, p=0.01, log_scale=TRUE)
  q_upper <- lpost_em$calc_lik_quantiles(U, p=0.99, log_scale=TRUE)

  dt_pred <- data.table(input = grid_info$input,
                        mean = pred$log_mean,
                        lower = q_lower,
                        upper = q_upper,
                        true = grid_info$lpost)

  # Design used to fit emulator.
  dt_design <- as.data.table(design_info)


  plt <- ggplot(dt_pred, aes(x=input)) +
            geom_line(aes(y=mean), color="blue") +
            geom_ribbon(aes(ymin=lower, ymax=upper), 
                        fill="skyblue", alpha=0.4) +
            geom_line(aes(y=true), color="black") +
            geom_point(aes(x=input, y=lpost), dt_design, 
                       color="red", inherit.aes=FALSE) +
            labs(xlab="u", ylab="log-posterior")
}


scale_lpost_var <- function(lpost_em, scale=1.0) {
  # Scales the marginal variance by multiplying by `scale`. The default of
  # 1.0 keeps the variance the same. Creates a deep copy of the emulator.
  lpost_em <- lpost_em$copy()
  new_var <- scale * lpost_em$emulator_model$gp_model$y1$nu_hat
  lpost_em$emulator_model$gp_model$y1$nu_hat <- new_var
  
  return(lpost_em)
}

```


```{r}
# Functions to compute normalized posterior density and normalized
# approximations of the posterior density. Assumes uniformly spaced
# grid.

normalize_density <- function(ldens_vals, du, log_scale=FALSE) {
  # Given unnormalized log-density values, returns the normalized
  # density, computed using LogSumExp. If `ldens_vals` is a matrix,
  # then each row is normalized. The normalizing constant is approximated
  # by the trapezoidal rule, and assumes an equally-spaced grid with 
  # spacing `du`.
  
  if(is.null(dim(ldens_vals))) ldens_vals <- matrix(ldens_vals, nrow=1L) 
  
  # Final column needs to be modified for the trapezoidal rule.
  ldens_vals_mod <- ldens_vals
  ldens_vals_mod[,ncol(ldens_vals_mod)] <- ldens_vals_mod[,ncol(ldens_vals_mod)] - log(2)
  log_sums <- matrixStats::rowLogSumExps(ldens_vals_mod)
  
  # Log of normalized density values.
  ldens_norm <- add_vec_to_mat_cols(-log_sums, ldens_vals) - log(du)
  
  if(log_scale) return(ldens_norm)
  return(exp(ldens_norm))
}

get_EP_dens_grid_1d <- function(lpost_em, input_grid, n_mc, log_scale=FALSE, ...) {
  
    assert_that(lpost_em$dim_input==1L)
    if(is.null(dim(input_grid))) input_grid <- matrix(input_grid, ncol=1L)
    n_grid <- nrow(input_grid)
    du <- abs(input_grid[2,] - input_grid[1,])
    
    # Simulate log-likelihood values. Return shape is (n_grid, n_mc) before
    # transposing.
    lpost_samp <- t(lpost_em$sample(input_grid, N_samp=n_mc, ...))
  
    # Normalize each trajectory.
    lpost_samp_norm <- normalize_density(lpost_samp, du, log_scale=TRUE)
    
    # Average density over trajectories.
    lpost_ep_norm <- matrixStats::colLogSumExps(lpost_samp_norm) - log(n_mc)
    
    if(log_scale) return(lpost_ep_norm)
    return(exp(lpost_ep_norm))
}


# Exact posterior density.
get_post_exact <- function(grid_info, log_scale=FALSE) {
  U <- drop(grid_info$input)
  du <- abs(U[2] - U[1])
  
  drop(normalize_density(grid_info$lpost, du, log_scale=log_scale))
}

# Expected likelihood approximation.
get_post_EL <- function(grid_info, lpost_em, log_scale=FALSE) {
  U <- matrix(grid_info$input, ncol=1L)
  pred <- lpost_em$predict_lik(U, return_var=FALSE, log_scale=TRUE)
  
  # Normalize density.
  U <- drop(U)
  du <- abs(U[2] - U[1])
  drop(normalize_density(pred$log_mean, du, log_scale=log_scale))
}


plot_post_comparison <- function(grid_info, design_info, lpost_em, 
                                 n_mc_EP=1e5, log_scale=FALSE, ...) {
  
  dt <- data.table(input = drop(grid_info$input),
                   exact = get_post_exact(grid_info, log_scale=log_scale),
                   EL = get_post_EL(grid_info, lpost_em, log_scale=log_scale),
                   EP = get_EP_dens_grid_1d(lpost_em, grid_info$input, n_mc=n_mc_EP, 
                                            log_scale=log_scale, ...))
  
  plt <- ggplot(dt, aes(x=input)) +
          geom_line(aes(y=exact), color="black") +
          geom_line(aes(y=EL), color="orange") +
          geom_line(aes(y=EP), color="green") +
          geom_vline(xintercept=design_info$input, color="red", 
                     linetype="dashed") +
          labs(x="u", y="posterior")
}

```







# Grid for plotting.
```{r}
n_grid <- 100L
p_grid <- .99
grid_min <- -3
grid_max <- 3

grid_info <- get_design(seq(grid_min, grid_max, length.out=n_grid))
grid_info_dt <- as.data.table(grid_info)
input_grid <- matrix(grid_info$input, ncol=1L)
```

```{r}
# Plot exact quantities.
plt_exact <- ggplot(grid_info_dt, aes(x=input)) +
              geom_line(aes(y=llik), color="blue") +
              geom_line(aes(y=lprior), color="red") +
              geom_line(aes(y=lpost), color="black")

plot(plt_exact)
```

# Log-likelihood emulation. 
```{r}
# Function to fit llik emulator.

fit_llik_em <- function(design) {
  # Fit GP.
  gp <- gpWrapperHet(matrix(design$input, ncol=1L), 
                     matrix(design$llik, ncol=1L),
                     scale_input=TRUE, normalize_output=TRUE)
  gp$set_gp_prior("Gaussian", "constant", include_noise=FALSE)
  gp$fit()
  
  # Create llikEmulator object. Shifting the mean so that the predictions
  # are for the log-posterior.
  lpost_em <- llikEmulatorGP("em_llik", gp, default_conditional=FALSE, 
                             default_normalize=TRUE, shift_func=lprior)
}

```


```{r}
# Design.

design_info <- get_design(qnorm(c(.1, .5, .9), sd=sqrt(sig2_prior)))

lpost_em <- fit_llik_em(design_info)
lpost_em2 <- scale_lpost_var(lpost_em, scale=2.0)
```

```{r}
# lpost em. 

plt <- plot_lpost_em(lpost_em, grid_info, design_info)

plot(plt)
```


```{r}
# Unnormalize posterior emulator (on log scale).

test <- lpost_em$predict(input_grid)
m <- test$mean + 0.5 * test$var
q99 <- test$mean + qnorm(.9999) * sqrt(test$var)


plt <- plot_post_em(lpost_em, grid_info, design_info)
plot(plt)
```


```{r}
plt <- plot_post_comparison(grid_info, design_info, lpost_em)
plt2 <- plot_post_comparison(grid_info, design_info, lpost_em2)

plot(plt)
plot(plt2)
```

```{r}
test <- get_post_EL(grid_info, lpost_em, log_scale=TRUE)

plot(drop(input_grid), test, type="l")

plot(drop(input_grid), exp(test), type="l")
```





