---
title: "noisy_mcmc_1d_noise_lvl_test"
output: html_document
---

Tests to do:
- Compare algs when noise level of GP is varied.
- Effect of stationary prior; test what happens with noisy MCMC.

Nice examples:

List 1:

List 2:
#2: Example where marginal is multimodal but mcwmh-joint is not (results in 
basically inflating the tails). mcwmh-ind has some divergence issues.
Design points: (-7, -2, 2, 7)

#6: This is an interesting one that illustrates a few different points.
Shows the benefit of the marginal approx over the mean approx. Interesting case
where the marginal sampler converges but the mcwmh samplers diverge. This makes
complete sense in this example: 



```{r}
set.seed(63462)

library(data.table)
library(ggplot2)
library(patchwork)
library(scales)
library(grid)
library(support)

base_dir <- file.path("/projectnb", "dietzelab", "arober", "bip-surrogates-paper")
code_dir <- file.path("/projectnb", "dietzelab", "arober", "gp-calibration")
src_dir <- file.path(code_dir, "src")

# Source required files.
source(file.path(src_dir, "general_helper_functions.r"))
source(file.path(src_dir, "statistical_helper_functions.r"))
source(file.path(src_dir, "plotting_helper_functions.r"))
source(file.path(src_dir, "mcmc_helper_functions.r"))
source(file.path(src_dir, "seq_design.r"))
source(file.path(src_dir, "gp_helper_functions.r"))
source(file.path(src_dir, "gpWrapper.r"))
source(file.path(src_dir, "llikEmulator.r"))
source(file.path(base_dir, "scripts", "helper", "sim_study_functions.r"))
```

# Target log density
```{r}
ldens <- function(x) dnorm(x, mean=0, sd=1, log=TRUE)
```

# Grid for plotting.
```{r}
n_grid <- 100L
grid_min <- -8
grid_max <- 8

grid_info <- list(input = seq(grid_min, grid_max, length.out=n_grid))
grid_info$ldens <- ldens(grid_info$input)
grid_info_dt <- as.data.table(grid_info)
```

```{r}
plt_ldens <- ggplot() +
              geom_line(aes(x=input, y=ldens), grid_info_dt, color="black")

plt_dens <- ggplot() +
              geom_line(aes(x=input, y=exp(ldens)), grid_info_dt, color="black")

plot(plt_ldens)
plot(plt_dens)
```

# Designs
```{r}
n_vals <- c(1, 2, 3, 4, 5)

# First design list is contained within [-2,2]. "Wide" design list is 
# contained within [-4,4].
design_list <- list()
design_list_wide <- list()

for(i in seq_along(n_vals)) {
  n <- n_vals[i]
  
  u <- c(-7, 7, seq(-2, 2, length.out=n))
  l <- ldens(u)
  design_list[[i]] <- list(input=u, ldens=l)
  
  u_wide <- c(-7, 7, seq(-4, 4, length.out=n))
  l_wide <- ldens(u_wide)
  design_list_wide[[i]] <- list(input=u_wide, ldens=l_wide)
}

design_list <- c(design_list, design_list_wide)
```


# Fit log-density emulator.
```{r}
em_list <- list()

for(i in seq_along(design_list)) {
  design <- design_list[[i]]
  
  # Fit GP.
  gp <- gpWrapperHet(matrix(design$input, ncol=1L), 
                     matrix(design$ldens, ncol=1L),
                     scale_input=TRUE, normalize_output=TRUE)
  gp$set_gp_prior("Gaussian", "constant", include_noise=FALSE)
  gp$fit()
  
  # Create llikEmulator object.
  em_list[[i]] <- llikEmulatorGP("em_ldens", gp, default_conditional=FALSE, 
                                 default_normalize=TRUE)
}
```


# Predict at grid points with log-density emulator.
```{r}
pred_list <- list()
U <- matrix(grid_info$input, ncol=1L)

for(i in seq_along(em_list)) {
  ldens_em <- em_list[[i]]
  ldens_pred <- ldens_em$predict(U)
  pred_list[[i]] <- list(ldens = em$predict(U, em_pred_list=ldens_pred),
                         dens = em$predict_lik(U, llik_pred_list=ldens_pred, log_scale=TRUE))
}
```


```{r}
# Plot ldens predictions.

idx <- 6L

CI_prob <- 0.9
q_lower <- 0.5 * (1-CI_prob)
alpha <- qnorm(q_lower)  
pred_dt <- as.data.table(pred_list[[idx]]$ldens)
pred_dt[, input := grid_info$input]
pred_dt[, upper := pred_dt$mean + alpha * sqrt(pred_dt$var)]
pred_dt[, lower := pred_dt$mean - alpha * sqrt(pred_dt$var)]

plt_ldens + 
  geom_point(aes(x=input, y=ldens), as.data.table(design_list[[idx]]), color="red") +
  geom_line(aes(x=input, y=mean), pred_dt, color="blue") +
  geom_ribbon(aes(x=input, ymin=lower, ymax=upper), pred_dt, fill="skyblue", alpha=0.4)

```

```{r}
# Plot dens predictions.
pred_dt <- cbind(pred_dt, as.data.table(pred_list[[idx]]$dens))
pred_dt[, upper_ln := qlnorm(1-q_lower, meanlog=mean, sdlog=sqrt(var))]
pred_dt[, lower_ln := qlnorm(q_lower, meanlog=mean, sdlog=sqrt(var))]

plt_dens + 
  geom_point(aes(x=input, y=exp(ldens)), as.data.table(design_list[[idx]]), color="red") +
  geom_line(aes(x=input, y=exp(log_mean)), pred_dt, color="blue") +
  geom_ribbon(aes(x=input, ymin=lower_ln, ymax=upper_ln), pred_dt, fill="skyblue", alpha=0.4) +
  theme_minimal()

```

# MCMC

```{r}
# MCMC Settings.

n_chain <- 4L
common_settings <- list(n_itr=10000L, try_parallel=TRUE, n_chain=n_chain,
                        itr_start=5000L, adapt_cov_prop=FALSE, 
                        adapt_scale_prop=FALSE, cov_prop=matrix(1),
                        log_scale_prop=0, 
                        par_init=matrix(seq(-2, 2, length.out=n_chain), ncol=1L))

mcmc_settings <- list(
  c(list(test_label="mean", approx_type="mean", adjustment="none",
         mcmc_func_name="mcmc_gp_unn_post_dens_approx"), common_settings),
  c(list(test_label="marginal", approx_type="marginal", adjustment="none",
         mcmc_func_name="mcmc_gp_unn_post_dens_approx"), common_settings),
  c(list(test_label="mcwmh-joint", mode="mcwmh", use_joint=TRUE, adjustment="none",
         mcmc_func_name="mcmc_noisy_llik"), common_settings),
  c(list(test_label="mcwmh-ind", mode="mcwmh", use_joint=FALSE, adjustment="none",
         mcmc_func_name="mcmc_noisy_llik"), common_settings),
  c(list(test_label="pm", mode="pseudo-marginal", use_joint=FALSE, adjustment="none",
         mcmc_func_name="mcmc_noisy_llik"), common_settings)
)

names(mcmc_settings) <- sapply(mcmc_settings, function(x) x$test_label)
```


```{r}
# Run MCMC.
mcmc_list <- list()
par_prior <- "flat"

for(i in seq_along(em_list)) {
  print(paste0("Emulator ", i))
  mcmc_list[[i]] <- run_mcmc_comparison(em_list[[i]], par_prior, 
                                        mcmc_settings, return=TRUE)
}


saveRDS(mcmc_list, "/projectnb/dietzelab/arober/bip-surrogates-paper/scripts/notebooks/mcmc_list_new.rds")
```

```{r}
plts <- get_trace_plots(mcmc_list[[6]]$marginal$samp)
for(plt in plts) plot(plt)
```

```{r}
samp_temp <- mcmc_list[[6]]$`mcwmh-joint`$samp
kde_plt <- get_1d_kde_plots(samp_temp)

plot(kde_plt[[1]]) + geom_line(aes(x=input, y=exp(ldens)), grid_info_dt, color="black")
```

```{r}
samp_temp <- mcmc_list[[6]]$`mcwmh-joint`$samp
kde_plt <- get_1d_kde_plots(samp_temp)

plot(kde_plt[[1]]) + geom_line(aes(x=input, y=exp(ldens)), grid_info_dt, color="black")
```









